Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-11-24 manuscript/manuscript.tex
Code cleanup for images
Refactored code for manuscript image production.
minor edits/fixes

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile channels in 2009, a
record monhtly high of 348,463 Service Requests in 
August 2020 (associated with the the COVID-19 
pandemic), the 2021 expansion to include the 
Metropolitan Transit Authority (MTA) NYC subway 
system (the biggest expansion in 311 history), 
and a yearly record of 3.23 million SRs in 2023.

@@

\item There are 41 columns (fields) of data for each row, with each
  row representing a single Service Request (SR). To facilitate 
  processing, column names were made lower case and 
  spaces were replaced with underscores, e.g. \texttt{complaint\_type}. 
  This naming convention is used throught this paper.

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]{SRs_by_Agency.pdf}
  	\caption{SR Counts by Agency for the time period 2022-2023}
	\label{fig:SRcountbyAgency}
\end{figure}

@@

\begin{figure}[tbp]
 \centering
  \includegraphics[width = \textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 Complaint Types for the time period 2022-2023} 
  \label{fig:SR_complaints}
\end{figure}

@@

\paragraph{Undocumented Fields:} The 311 SR 
The 311 Service Request Data Dictionary identifies 
41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded if manually selected. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' and include:

@@

While one can likely infer the meaning of these ``computed'' 
fields, uncertainty remains as to their derivation and 
usage. These six fields are not included in
the 311 Service Request Data Dictionary. Additionally, our analysis 
revealed that these six ``computed'' fields have 
significant data validity issues, leading us to categorize these six fields 
as experimental and \textit{not for public use}. An update 
to the Data Dictionary would seem appropriate. Accordingly, 
we will not address these fields in our analysis.

@@

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=\textwidth]{BlankFields.pdf}
	\caption{Number and Percentage of Empty/Blank Entries for the time period 2022-2023}
	\label{fig:blank_fields}
\end{figure}

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. For example, the \texttt{latitude} 
and \texttt{longitude} fields were found to all fall within the 
geographic boundaries of New York City. Likewise, 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the 311 Service Request Data Dictionary specifies very few domains of 
acceptable values for other fields. Nonetheless, the following fields were tested and found to comply with their expected domains, as determined by public 
usage and as evidenced in other historical datasets:

@@

\begin{figure}[tbp]
  \centering
  % First figure
  \begin{subfigure}[t]{0.495\textwidth} % Adjust width as needed
    \centering
    \includegraphics[width=\textwidth]{post_closed_violin_chart.pdf}
    \caption{Post-closed \texttt{resolution\_action\_update\_date}(s) >30 days}
    \label{fig:resolution-violin}
  \end{subfigure}
  \hfill % Horizontal space between subfigures
  % Second figure
  \begin{subfigure}[t]{0.495\textwidth} % Adjust width as needed
    \centering
    \includegraphics[width=\textwidth]{negative_duration_SR_violin.pdf}
    \caption{Distribution of Negative Durations.}
    \label{fig:negative-duration-violin}
  \end{subfigure}
  % Main figure caption
  \caption{Comparison of Violin Plots: (a) Post-Closed Resolution and (b) Negative Durations for the time period 2022-2023}
  \label{fig:violin-plots}
\end{figure}

@@

\begin{table}[tbp]
  \centering
  \caption{Sample  Negative Durations [excluding extreme values that are <$-$730 days]}
   \label{tab:largest-errors}
 	\begin{tabular}{l l l r l}
    \toprule
    \texttt{created\_date} & \texttt{closed\_date} & {Duration (days)}  & {Agency} \\
    \midrule
    2023-01-27 14:40:00 & 2022-01-14 14:40:00 & $-$378 & DOT \\
    2023-01-18 10:06:00 & 2022-01-12 10:06:00 & $-$371 & DOT \\
    2023-01-27 14:36:00 & 2022-01-22 14:35:00 & $-$370 & DOT \\
    2023-01-11 11:10:00 & 2022-01-09 11:10:00 & $-$367 & DOT \\
    2023-12-18 03:13:00 & 2023-01-16 13:10:00 & $-$335 & DOT \\
    \bottomrule
    \end{tabular}
 \end{table}

@@

\paragraph{Created or Closed at Midnight:}
Another issue involves the unusually large number of 
SRs where the \texttt{created\_date} and \texttt{closed\_date} 
indicate ``created'' or ``closed'' exactly at midnight (00:00:00), 
to the second. Normally, SR creation and closure follows 
the work-day schedule, with most SRs created 
during working hours and fewer at night or in the 
early morning. However, there is a significantly higher 
number of SRs closed exactly at midnight, as well 
as a large number created exactly at midnight. The magnitude 
of this anomaly is well outside the 3$\sigma$ level as shown 
in Figure~\ref{fig:stacked}

\begin{figure}[tbp]
    \centering
    % First subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs Created Exactly on the Hour}
        \label{fig:busiestcreated}
    \end{subfigure}
    \hfill % Add horizontal space between subfigures
    % Second subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs Closed Exactly on the Hour}
        \label{fig:busiestclosed}
    \end{subfigure}
    % Main caption
    \caption{Comparison of SRs Created \& Closed by Hour-of-the-Day for the time period 2022-2023}
    \label{fig:comparison}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{post_closed_violin_chart.pdf}
  \caption{Post-closed \texttt{resolution\_action\_update\_date}(s) >30 days.}
  \label{fig:resolution-violin}
\end{figure}

@@

\paragraph{\texttt{landmark} \& \texttt{street\_name}:} Per the 
311 Service Request Data Dictionary, the \texttt{landmark} 
field \textit{can refer to any noteworthy location, 
including but not limited to, parks, hospitals, airports, sports facilities, 
performance spaces, etc..} However, most entries are actually 
street names, with a 62\% match to \texttt{street\_name} field. 
Even many of the non-matches (excluding blanks) appear to be 
a logical match, differing only in spelling or formatting 
(e.g., ``NINTH AVE'' vs. ``9 AVE''). This indicates a possible 
misuse of the \texttt{landmark} field as well as a high level of redundancy.

@@

The 311 Service Request Data Dictionary would greatly benefit from an update. We found several discrepancies between the Data Dictionary and the actual data, which could potentially mislead analytic efforts. Here are some examples:

>>>>

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile channels in 2009, a
record monhtly high of 348,463 SRs in 
August 2020 (associated with the the COVID-19 
pandemic), the 2021 expansion to include the 
Metropolitan Transit Authority (MTA) NYC subway 
system (the biggest expansion in 311 history), 
and a yearly record of 3.23 million SRs in 2023.

@@

\item There are 41 columns (fields) of data for each row, with each
  row representing a single SR. To facilitate 
  processing, column names were made lower case and 
  spaces were replaced with underscores, e.g. \texttt{complaint\_type}. 
  This naming convention is used throught this paper.

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]{SRs_by_Agency-crop.pdf}
  	\caption{311 SR counts by agency for the time period 2022--2023.}
	\label{fig:SRcountbyAgency}
\end{figure}

@@

\begin{figure}[tbp]
 \centering
  \includegraphics[width = \textwidth]{SR_by_Complaint_Type-crop.pdf} 
  \caption{Top 20 complaint types for the 311 SR during 2022--2023.} 
  \label{fig:SR_complaints}
\end{figure}

@@

\paragraph{Undocumented Fields:} The 311 SR 
The 311 SR Data Dictionary identifies 
41 data columns (fields) and provides related information 
for each. However, there are six additional fields that can be
optionally downloaded if manually selected. These fields are 
visible in the portal's ``Column manager'' widget, labeled with the prefix 
``@computed\_region.'' and include:

@@

While one can likely infer the meaning of these ``computed'' 
fields, uncertainty remains as to their derivation and 
usage. These six fields are not included in
the 311 SR Data Dictionary. Additionally, our analysis 
revealed that these six ``computed'' fields have 
significant data validity issues, leading us to categorize these six fields 
as experimental and \textit{not for public use}. An update 
to the Data Dictionary would seem appropriate. Accordingly, 
we will not address these fields in our analysis.

@@

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=\textwidth]{BlankFields-cro.pdf}
	\caption{Number and percentage of empty/blank entries for the
          311 SR data during 2022--2023.}
	\label{fig:blank_fields}
\end{figure}

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and isolated from the analysis. For example, the \texttt{latitude} 
and \texttt{longitude} fields were found to all fall within the 
geographic boundaries of New York City. Likewise, 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the 311 SR Data Dictionary specifies very few domains
of acceptable values for other fields. Nonetheless, the following
fields were tested and found to comply with their expected
domains, as determined by public
usage and as evidenced in other historical datasets:

@@

\begin{figure}[tbp]
  \centering
  % First figure
  \begin{subfigure}[t]{0.495\textwidth} % Adjust width as needed
    \centering
    \includegraphics[width=\textwidth]{negative_duration_SR_violin.pdf}
    \caption{Distribution of negative response times.}
    \label{fig:negative-duration-violin}
  \end{subfigure}
  \hfill % Horizontal space between subfigures
  % Second figure
  \begin{subfigure}[t]{0.495\textwidth} % Adjust width as needed
    \centering
    \includegraphics[width=\textwidth]{post_closed_violin_chart.pdf}
    \caption{Post-closed \texttt{resolution\_action\_update\_date}(s) >30 days}
    \label{fig:resolution-violin}
  \end{subfigure}
  % Main figure caption
  \caption{Violin plots for negative response times and post-closed
    resolution during 2022--2023.}
  \label{fig:violin-plots}
\end{figure}

@@

\begin{table}[tbp]
  \centering
  \caption{Sample  negative durations [excluding extreme values that are <$-$730 days]}
   \label{tab:largest-errors}
 	\begin{tabular}{l l l r l}
    \toprule
    \texttt{created\_date} & \texttt{closed\_date} & {Duration (days)}  & {Agency} \\
    \midrule
    2023-01-27 14:40:00 & 2022-01-14 14:40:00 & $-$378 & DOT \\
    2023-01-18 10:06:00 & 2022-01-12 10:06:00 & $-$371 & DOT \\
    2023-01-27 14:36:00 & 2022-01-22 14:35:00 & $-$370 & DOT \\
    2023-01-11 11:10:00 & 2022-01-09 11:10:00 & $-$367 & DOT \\
    2023-12-18 03:13:00 & 2023-01-16 13:10:00 & $-$335 & DOT \\
    \bottomrule
    \end{tabular}
 \end{table}

@@

\paragraph{Created or Closed at Midnight:}
Another issue involves the unusually large number of 
SRs where the \texttt{created\_date} and \texttt{closed\_date} 
indicate ``created'' or ``closed'' exactly at midnight (00:00:00), 
to the second. Normally, SR creation and closure follows 
the work-day schedule, with most SRs created 
during working hours and fewer at night or in the 
early morning. However, there is a significantly higher 
number of SRs closed exactly at midnight, as well 
as a large number created exactly at midnight. The magnitude 
of this anomaly is well outside the 3$\sigma$ level as shown 
in Figure~\ref{fig:exacthours}

\begin{figure}[tbp]
    \centering
    % First subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs created exactly on the hour}
        \label{fig:busiestcreated}
    \end{subfigure}
    \hfill % Add horizontal space between subfigures
    % Second subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs closed exactly on the hour}
        \label{fig:busiestclosed}
    \end{subfigure}
    % Main caption
    \caption{Comparison of 311 SRs created \& closed by hour-of-the-day during 2022--2023.}
    \label{fig:exacthours}
\end{figure}

@@

@@

\paragraph{\texttt{landmark} \& \texttt{street\_name}:} Per the 
311 SR Data Dictionary, the \texttt{landmark} 
field \textit{can refer to any noteworthy location, 
including but not limited to, parks, hospitals, airports, sports facilities, 
performance spaces, etc..} However, most entries are actually 
street names, with a 62\% match to \texttt{street\_name} field. 
Even many of the non-matches (excluding blanks) appear to be 
a logical match, differing only in spelling or formatting 
(e.g., ``NINTH AVE'' vs. ``9 AVE''). This indicates a possible 
misuse of the \texttt{landmark} field as well as a high level of redundancy.

@@

The 311 SR Data Dictionary would greatly benefit from an update. We
found several discrepancies between the Data Dictionary and the actual
data, which could potentially mislead analytic efforts. Here are some
examples:
